---
title: "U.S. Housing Affordability, 1993–2024"
subtitle: "STAT 184 · Section 4 Course Project"
author: "Paige · Tessa · Kunal · Sam"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
    theme: flatly
    highlight: tango
    code_folding: hide
    fig_width: 9
    fig_height: 5
    df_print: kable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

---

## 1 · Introduction

Over the past three decades, American households have faced a widening gap between what homes cost and what they earn. This report quantifies that gap using four publicly available datasets downloaded directly from the **Federal Reserve Economic Data (FRED)** system operated by the Federal Reserve Bank of St. Louis.

We examine eight metrics:

| # | Metric | FRED Series ID | Frequency |
|---|--------|---------------|-----------|
| 1 | Median home sales price | `MSPUS` | Quarterly |
| 2 | Estimated median monthly rent | `CUUR0000SEHA` | Monthly |
| 3 | Price-to-rent ratio | Derived | Annual |
| 4 & 5 | Median household income — nominal & real | `MEHOINUSA646N` | Annual |
| 5 | Nominal & real income growth (YoY %) | Derived | Annual |
| 6 | CPI-U all items & annual inflation rate | `CPIAUCSL` | Monthly |
| 7 | Price-to-income ratio | Derived | Annual |
| 8 | All metrics indexed to year 2000 | Derived | Annual |

**Data files** (`data/`) were downloaded as CSV directly from:
`https://fred.stlouisfed.org/graph/fredgraph.csv?id=<SERIES_ID>`

No API calls or web scraping are performed at analysis time — the CSVs are read from disk.

---

## 2 · Setup

```{r libraries}
library(readr)
library(dplyr)
library(tidyr)
library(lubridate)
library(ggplot2)
library(scales)
library(patchwork)
library(knitr)
```

```{r theme-palette}
theme_housing <- function(base = 13) {
  theme_minimal(base_size = base) +
    theme(
      plot.title       = element_text(face = "bold", size = base + 1,
                                      margin = margin(b = 5)),
      plot.subtitle    = element_text(color = "grey40", size = base - 1),
      plot.caption     = element_text(color = "grey55", size = 8.5,
                                      hjust = 0, margin = margin(t = 8)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "grey92"),
      axis.title       = element_text(size = base - 1),
      legend.position  = "bottom",
      legend.title     = element_blank()
    )
}

CAP  <- paste0(
  "Sources: FRED — MSPUS, CUUR0000SEHA, MEHOINUSA646N, CPIAUCSL",
  "\nSTAT 184 Section 4 · Paige, Tessa, Kunal, Sam"
)
XBRK <- seq(1993, 2024, by = 3)

BLUE  <- "#1B4F8A"; RUST  <- "#C1440E"; GREEN <- "#1F7A4B"
PURP  <- "#6B3FA0"; GOLD  <- "#B8860B"; TEAL  <- "#1A7A7A"
```

---

## 3 · Data Ingestion & Wrangling

```{r load-data}
# ── Median home price (quarterly) ────────────────────────────────────────────
home_q <- read_csv("data/MSPUS.csv", show_col_types = FALSE) %>%
  rename(date = observation_date, home_price = MSPUS) %>%
  mutate(date = ymd(date), year = year(date))

home_annual <- home_q %>%
  group_by(year) %>%
  summarise(median_home_price = mean(home_price, na.rm = TRUE), .groups = "drop")

# ── CPI rent component (monthly index, 1982-84 = 100) ────────────────────────
# Calibrated to U.S. Census Bureau 2000 median gross rent: $602/month
rent_raw <- read_csv("data/CUUR0000SEHA.csv", show_col_types = FALSE) %>%
  rename(date = observation_date, rent_idx = CUUR0000SEHA) %>%
  mutate(date = ymd(date), year = year(date))

rent_annual <- rent_raw %>%
  group_by(year) %>%
  summarise(rent_idx = mean(rent_idx, na.rm = TRUE), .groups = "drop")

base_idx_2000 <- rent_annual %>% filter(year == 2000) %>% pull(rent_idx)

rent_annual <- rent_annual %>%
  mutate(median_rent = rent_idx / base_idx_2000 * 602) %>%
  select(year, median_rent)

# ── Median household income (nominal, annual) ─────────────────────────────────
income_raw <- read_csv("data/MEHOINUSA646N.csv", show_col_types = FALSE) %>%
  rename(date = observation_date, median_hh_income_nom = MEHOINUSA646N) %>%
  mutate(year = year(ymd(date))) %>%
  select(year, median_hh_income_nom)

# ── CPI-U all items (monthly → annual average) ────────────────────────────────
cpi_raw <- read_csv("data/CPIAUCSL.csv", show_col_types = FALSE) %>%
  rename(date = observation_date, cpi = CPIAUCSL) %>%
  mutate(date = ymd(date), year = year(date))

cpi_annual <- cpi_raw %>%
  group_by(year) %>%
  summarise(cpi = mean(cpi, na.rm = TRUE), .groups = "drop")

# ── Master data frame with derived metrics ────────────────────────────────────
cpi_2024 <- cpi_annual %>% filter(year == 2024) %>% pull(cpi)

housing <- home_annual %>%
  left_join(rent_annual,  by = "year") %>%
  left_join(income_raw,   by = "year") %>%
  left_join(cpi_annual,   by = "year") %>%
  filter(year >= 1993, year <= 2024) %>%
  arrange(year) %>%
  mutate(
    # Inflation-adjusted income (2024 dollars)
    median_hh_income_real = median_hh_income_nom * (cpi_2024 / cpi),
    # Price-to-rent ratio (annual price ÷ annualised monthly rent)
    price_to_rent         = median_home_price / (median_rent * 12),
    # Price-to-income ratio
    price_to_income       = median_home_price / median_hh_income_nom,
    # Year-over-year growth rates
    income_growth_nom     = (median_hh_income_nom  / lag(median_hh_income_nom)  - 1) * 100,
    income_growth_real    = (median_hh_income_real / lag(median_hh_income_real) - 1) * 100,
    cpi_yoy               = (cpi / lag(cpi) - 1) * 100
  )
```

### 3.1 · Data Summary

```{r summary-table}
housing %>%
  filter(year %in% c(1993, 2000, 2006, 2012, 2019, 2022, 2024)) %>%
  transmute(
    Year                  = year,
    `Median Home Price`   = dollar(round(median_home_price, -2)),
    `Est. Monthly Rent`   = dollar(round(median_rent)),
    `HH Income (Nominal)` = dollar(median_hh_income_nom),
    `HH Income (Real)`    = dollar(round(median_hh_income_real)),
    `Price/Rent`          = round(price_to_rent, 1),
    `Price/Income`        = round(price_to_income, 2),
    `CPI`                 = round(cpi, 1),
    `Inflation %`         = ifelse(is.na(cpi_yoy), "—", paste0(round(cpi_yoy, 1), "%"))
  ) %>%
  kable(caption = "Key housing metrics at selected benchmark years",
        align   = c("c","r","r","r","r","r","r","r","r"))
```

---

## 4 · Results

### Figure 1 · Median Home Sales Price

```{r fig1-home-price, fig.cap="Median U.S. home sales price in nominal dollars. Shaded bands mark the 2008 housing crisis and the 2020–2022 pandemic surge."}
ggplot(housing, aes(x = year, y = median_home_price)) +
  annotate("rect", xmin = 2007.5, xmax = 2011.5,
           ymin = -Inf, ymax = Inf, fill = "grey70", alpha = 0.15) +
  annotate("text", x = 2009.5, y = 510000, label = "Housing\ncrisis",
           color = "grey45", size = 3, hjust = 0.5) +
  annotate("rect", xmin = 2019.8, xmax = 2022.8,
           ymin = -Inf, ymax = Inf, fill = "#FFF3CD", alpha = 0.4) +
  annotate("text", x = 2021.3, y = 510000, label = "Pandemic\nboom",
           color = "#856404", size = 3, hjust = 0.5) +
  geom_area(fill = BLUE, alpha = 0.12) +
  geom_line(color = BLUE, linewidth = 1.1) +
  geom_point(color = BLUE, size = 2, shape = 21, fill = "white", stroke = 1.2) +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "K"),
                     limits = c(0, 560000)) +
  labs(title    = "Median U.S. Home Sales Price (1993–2024)",
       subtitle = "Nominal dollars · FRED: MSPUS (quarterly, annual average shown)",
       x = "Year", y = "Median Sales Price", caption = CAP) +
  theme_housing()
```

From **\$151K in 1993** to a peak of **\$486K in 2022**, median home prices rose **217%** in nominal terms over the study period — falling slightly to **\$477K by 2024** after the Federal Reserve's rate-hiking cycle. Two inflection points stand out: the 2006–2011 bust (prices fell ~35% from peak to trough) and the 2020–2022 pandemic surge, which was faster and larger than the mid-2000s bubble in percentage terms.

---

### Figure 2 · Estimated Median Monthly Rent

```{r fig2-rent, fig.cap="Estimated median U.S. gross monthly rent. The CPI rent index is calibrated to the U.S. Census Bureau 2000 median of $602/month."}
ggplot(housing, aes(x = year, y = median_rent)) +
  geom_area(fill = RUST, alpha = 0.12) +
  geom_line(color = RUST, linewidth = 1.1) +
  geom_point(color = RUST, size = 2, shape = 21, fill = "white", stroke = 1.2) +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(labels = label_dollar(accuracy = 1)) +
  labs(title    = "Estimated Median U.S. Monthly Rent (1993–2024)",
       subtitle = "CPI rent-of-primary-residence index (CUUR0000SEHA) · anchored to Census 2000 median ($602/mo)",
       x = "Year", y = "Estimated Monthly Rent", caption = CAP) +
  theme_housing()
```

Unlike home prices, rent growth has been **monotonically upward** — no bust, no correction. From an estimated **\$455/month in 1993** to **\$1,876/month in 2024**, rents rose **312%** nominally, outpacing both general CPI (+117%) and nominal income growth (+158%) over the same period. The steepest acceleration occurred after 2019, driven by supply shortages and post-pandemic demand shifts.

> **Methodology note:** A direct long-run median rent series is not available on FRED. We use the BLS CPI rent-of-primary-residence component (`CUUR0000SEHA`) as an index, anchored to the U.S. Census Bureau's 2000 American Housing Survey median gross rent of \$602/month. This gives an internally consistent trend but will diverge from spot-market rents in high-cost metros.

---

### Figure 3 · Price-to-Rent Ratio

```{r fig3-ptr, fig.cap="The price-to-rent ratio compares the annual cost of owning versus the annual cost of renting. Conventional thresholds: below 15 favors buying; above 20 favors renting."}
ptr_zones <- data.frame(
  ymin = c(0, 15, 20), ymax = c(15, 20, 50),
  fill = c("#d4edda", "#fff3cd", "#f8d7da")
)

ggplot() +
  geom_rect(data = ptr_zones,
            aes(xmin = -Inf, xmax = Inf, ymin = ymin, ymax = ymax),
            fill = ptr_zones$fill, alpha = 0.45) +
  geom_line(data = housing, aes(x = year, y = price_to_rent),
            color = GOLD, linewidth = 1.3) +
  geom_point(data = housing, aes(x = year, y = price_to_rent),
             color = GOLD, size = 2.3) +
  annotate("text", x = 1994, y =  7.5, hjust = 0, color = "#155724",
           size = 3.1, label = "Buy favored  (ratio < 15)") +
  annotate("text", x = 1994, y = 17.3, hjust = 0, color = "#856404",
           size = 3.1, label = "Neutral  (15–20)") +
  annotate("text", x = 1994, y = 22.8, hjust = 0, color = "#721c24",
           size = 3.1, label = "Rent favored  (ratio > 20)") +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(breaks = seq(0, 50, by = 5),
                     limits = c(0, 50)) +
  labs(title    = "Price-to-Rent Ratio (1993–2024)",
       subtitle = "Annual median home price ÷ (estimated monthly rent × 12)",
       x = "Year", y = "Price-to-Rent Ratio", caption = CAP) +
  theme_housing()
```

The price-to-rent ratio has remained **entirely in "rent-favored" territory** throughout the entire 1993–2024 window, ranging from **20.2 to 39.9**. It peaked in the mid-2000s bubble (~40×), fell to ~22× post-crisis, recovered during the 2010s, then spiked again during the 2020–2022 pandemic surge before easing to ~21× by 2024. By this metric, renting has been the financially rational near-term choice for a typical U.S. household for over three decades — though the calculus varies significantly by local market and holding period.

---

### Figure 4 · Median Household Income: Nominal vs. Real

```{r fig4-income, fig.cap="Nominal income in current dollars vs. real income adjusted to 2024 dollars using CPI-U."}
income_long <- housing %>%
  select(year, median_hh_income_nom, median_hh_income_real) %>%
  pivot_longer(-year, names_to = "type", values_to = "income") %>%
  mutate(type = recode(type,
    median_hh_income_nom  = "Nominal",
    median_hh_income_real = "Real (2024 $)"))

ggplot(income_long, aes(x = year, y = income, color = type, linetype = type)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2) +
  scale_color_manual(values    = c("Nominal" = GREEN, "Real (2024 $)" = "#0A4F2E")) +
  scale_linetype_manual(values = c("Nominal" = "solid", "Real (2024 $)" = "dashed")) +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(labels = label_dollar(scale = 1e-3, suffix = "K")) +
  labs(title    = "Median U.S. Household Income: Nominal vs. Real (1993–2024)",
       subtitle = "Real income deflated to 2024 dollars using annual average CPI-U (CPIAUCSL)",
       x = "Year", y = "Median Household Income", caption = CAP) +
  theme_housing()
```

Nominal income grew from **\$31,241 in 1993** to **\$80,610 in 2024** (+158%). But after adjusting for inflation, **real income grew only 19%** over the same 31 years — from \$67,836 to \$80,610 in 2024 dollars. Real incomes peaked around 2019 (\$84,297), then declined sharply during the 2021–2022 inflation surge before recovering. The gap between the solid (nominal) and dashed (real) lines widens over time, reflecting the cumulative erosion of purchasing power.

---

### Figure 5 · Nominal & Real Income Growth (YoY %)

```{r fig5-income-growth, fig.cap="Year-over-year percent change in nominal vs. real (inflation-adjusted) median household income."}
growth_long <- housing %>%
  filter(!is.na(income_growth_nom)) %>%
  select(year, income_growth_nom, income_growth_real) %>%
  pivot_longer(-year, names_to = "type", values_to = "pct") %>%
  mutate(type = recode(type,
    income_growth_nom  = "Nominal",
    income_growth_real = "Real (2024 $)"))

ggplot(growth_long, aes(x = year, y = pct, fill = type)) +
  geom_col(position = "dodge", width = 0.75, alpha = 0.85) +
  geom_hline(yintercept = 0, linewidth = 0.5, color = "grey30") +
  scale_fill_manual(values = c("Nominal" = GREEN, "Real (2024 $)" = "#A8D5B5")) +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 0.1)) +
  labs(title    = "Year-over-Year Household Income Growth: Nominal vs. Real (1994–2024)",
       subtitle = "Negative real growth bars signal a net loss of purchasing power",
       x = "Year", y = "YoY Growth (%)", caption = CAP) +
  theme_housing()
```

Nominal income peaked at **+8.7% YoY in 2023** (a pandemic catch-up effect) but this was largely offset by concurrent inflation. Real income turned **sharply negative in 2021–2022** as the CPI surge (+8% peak) outpaced wage growth — the worst real income decline since the early 1980s. Years 2001, 2008, and 2020 also show real income contraction aligned with recessions.

---

### Figure 6 · CPI Level & Annual Inflation Rate

```{r fig6-cpi, fig.height=8, fig.cap="Top: CPI-U all-items index (1982-84=100). Bottom: Annual inflation rate with the Fed's 2% target shown."}
p6a <- ggplot(housing, aes(x = year, y = cpi)) +
  geom_area(fill = PURP, alpha = 0.12) +
  geom_line(color = PURP, linewidth = 1.1) +
  scale_x_continuous(breaks = XBRK) +
  labs(title    = "CPI-U All Items Index (1993–2024)",
       subtitle = "Annual average · FRED: CPIAUCSL (base: 1982–84 = 100)",
       x = NULL, y = "CPI Index") +
  theme_housing() +
  theme(plot.caption   = element_blank(),
        axis.text.x    = element_blank(),
        axis.ticks.x   = element_blank())

p6b <- housing %>%
  filter(!is.na(cpi_yoy)) %>%
  ggplot(aes(x = year, y = cpi_yoy, fill = cpi_yoy > 2)) +
  geom_col(alpha = 0.85) +
  geom_hline(yintercept = 2, linetype = "dashed",
             color = "grey40", linewidth = 0.7) +
  annotate("text", x = 1994.5, y = 2.35, hjust = 0,
           label = "Fed 2% target", color = "grey40", size = 3) +
  scale_fill_manual(values = c("TRUE" = PURP, "FALSE" = "#C0392B"),
                    guide = "none") +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(labels = label_percent(scale = 1, accuracy = 0.1)) +
  labs(title    = NULL,
       subtitle = "Annual CPI-U inflation rate (%) · purple = above 2% target",
       x = "Year", y = "Inflation (%)", caption = CAP) +
  theme_housing()

p6a / p6b + plot_layout(heights = c(1, 1))
```

The CPI-U index rose from **144.5 in 1993** to **313.7 in 2024** — meaning the same basket of goods costs **117% more** than it did 31 years ago. Inflation remained close to or below the Fed's 2% target from 1997 through 2020, then surged to **7.96% in 2022** — the highest annual rate in the dataset — driven by pandemic-related supply chain disruptions, energy prices, and excess demand. The 2022 spike had a significant compounding effect on real wages and housing affordability.

---

### Figure 7 · Price-to-Income Ratio

```{r fig7-pti, fig.cap="The price-to-income ratio measures housing affordability. A ratio of 3× is the historical U.S. norm; 5× is widely cited as a stress threshold."}
ggplot(housing, aes(x = year, y = price_to_income)) +
  annotate("rect", xmin = -Inf, xmax = Inf, ymin = 5, ymax = Inf,
           fill = "#f8d7da", alpha = 0.25) +
  geom_area(fill = TEAL, alpha = 0.10) +
  geom_line(color = TEAL, linewidth = 1.3) +
  geom_point(color = TEAL, size = 2.2) +
  geom_hline(yintercept = 3, linetype = "dashed",
             color = "grey45", linewidth = 0.8) +
  geom_hline(yintercept = 5, linetype = "dotted",
             color = "#C0392B", linewidth = 0.8) +
  annotate("text", x = 1994, y = 3.18, hjust = 0, color = "grey40",
           size = 3, label = "Historical norm (~3×)") +
  annotate("text", x = 1994, y = 5.2, hjust = 0, color = "#C0392B",
           size = 3, label = "Affordability stress threshold (~5×)") +
  scale_x_continuous(breaks = XBRK) +
  scale_y_continuous(labels = label_number(suffix = "×", accuracy = 0.1),
                     limits = c(0, 9)) +
  labs(title    = "Price-to-Income Ratio (1993–2024)",
       subtitle = "Median home sales price ÷ median household income (nominal)",
       x = "Year", y = "Price-to-Income Ratio", caption = CAP) +
  theme_housing()
```

The price-to-income ratio has **never been at or below the historical 3× norm** in this dataset — it was already at **4.8× in 1993**. It peaked at **7.65× in 2006** during the housing bubble, fell to **4.97× in 2012** post-crisis, then climbed steadily to **6.51× by 2022** before easing slightly to **5.92× in 2024**. The entire 2019–2024 period sits above the affordability stress threshold of 5×, indicating that housing has been severely unaffordable for the median U.S. household for at least six consecutive years.

---

### Figure 8 · All Metrics Indexed to Year 2000

```{r fig8-indexed, fig.height=6, fig.cap="All five metrics set to 100 in the year 2000. Divergence above the baseline reveals relative affordability erosion."}
idx_df <- housing %>%
  filter(year >= 2000) %>%
  mutate(across(
    c(median_home_price, median_rent,
      median_hh_income_nom, median_hh_income_real, cpi),
    ~ . / first(.) * 100
  )) %>%
  select(year, median_home_price, median_rent,
         median_hh_income_nom, median_hh_income_real, cpi) %>%
  pivot_longer(-year, names_to = "metric", values_to = "idx") %>%
  mutate(metric = recode(metric,
    median_home_price     = "Home Price",
    median_rent           = "Rent",
    median_hh_income_nom  = "Nominal Income",
    median_hh_income_real = "Real Income",
    cpi                   = "CPI"))

palette8 <- c("Home Price"     = BLUE,  "Rent"           = RUST,
              "Nominal Income" = GREEN, "Real Income"    = "#0A4F2E",
              "CPI"            = PURP)
ltype8   <- c("Home Price" = "solid", "Rent"          = "solid",
              "Nominal Income" = "solid", "Real Income" = "dashed",
              "CPI"            = "solid")

ggplot(idx_df, aes(x = year, y = idx, color = metric, linetype = metric)) +
  geom_hline(yintercept = 100, color = "grey70", linetype = "dotted") +
  geom_line(linewidth = 1.15) +
  annotate("text", x = 2001, y = 103.5, hjust = 0, color = "grey55",
           size = 2.9, label = "Baseline = 100 (year 2000)") +
  scale_color_manual(values = palette8) +
  scale_linetype_manual(values = ltype8) +
  scale_x_continuous(breaks = seq(2000, 2024, by = 4)) +
  scale_y_continuous(breaks = seq(50, 350, by = 50)) +
  guides(color    = guide_legend(nrow = 2),
         linetype = guide_legend(nrow = 2)) +
  labs(title    = "Housing Costs, Income & Inflation — Indexed to Year 2000",
       subtitle = "All series = 100 in 2000 · wider gap = greater affordability pressure",
       x = "Year", y = "Index (2000 = 100)", caption = CAP) +
  theme_housing()
```

This panel tells the central story of the analysis. From 2000 to 2024:

- **Home prices** rose to index **210** (up 110%)
- **Rent** rose to index **312** (up 212%)
- **Nominal income** rose to index **192** (up 92%)
- **CPI** rose to index **182** (up 82%)
- **Real income** barely moved — index only **105** (up just 5%)

The gap between home prices and real income has grown from zero in 2000 to **105 index points by 2024**, and the rent-income gap is even wider. Every dollar of real income growth has been consumed many times over by housing cost growth.

---

## 5 · Key Findings

```{r findings-table, echo=FALSE}
findings <- data.frame(
  Metric       = c("Median Home Price", "Estimated Rent", "Nominal Income",
                   "Real Income", "CPI", "Price-to-Rent (2024)",
                   "Price-to-Income (2024)", "Peak Inflation"),
  `1993 Value` = c("$151K", "$455/mo", "$31,241", "$67,836 (2024$)",
                   "144.5", "—", "—", "—"),
  `2024 Value` = c("$477K", "$1,876/mo", "$80,610", "$80,610 (2024$)",
                   "313.7", "21.2×", "5.92×", "7.96% (2022)"),
  `Change`     = c("+217%", "+312%", "+158%", "+19% (real)",
                   "+117%", "Rent-favored entire period",
                   "Above 5× stress threshold since 2019",
                   "Highest in 40 years"),
  check.names  = FALSE
)
kable(findings, caption = "Summary of key findings across all eight metrics")
```

**Core conclusion:** Median home prices rose 217% and rents rose 312% between 1993 and 2024, while real (inflation-adjusted) household income grew only 19%. The price-to-income ratio has remained above the 5× affordability stress threshold since 2019, and the price-to-rent ratio has been in "rent-favored" territory for the entire study period. The 2021–2022 inflation spike — the worst in four decades — further compressed real purchasing power at exactly the moment housing costs peaked, creating the most severe affordability conditions in the dataset.

---

## 6 · Data Sources

| Series ID | Description | Source | URL |
|-----------|-------------|--------|-----|
| `MSPUS` | Median Sales Price of Houses Sold | U.S. Census Bureau / HUD via FRED | [fred.stlouisfed.org/series/MSPUS](https://fred.stlouisfed.org/series/MSPUS) |
| `CUUR0000SEHA` | CPI: Rent of Primary Residence | U.S. BLS via FRED | [fred.stlouisfed.org/series/CUUR0000SEHA](https://fred.stlouisfed.org/series/CUUR0000SEHA) |
| `MEHOINUSA646N` | Real Median Household Income | U.S. Census Bureau via FRED | [fred.stlouisfed.org/series/MEHOINUSA646N](https://fred.stlouisfed.org/series/MEHOINUSA646N) |
| `CPIAUCSL` | CPI-U All Items | U.S. BLS via FRED | [fred.stlouisfed.org/series/CPIAUCSL](https://fred.stlouisfed.org/series/CPIAUCSL) |
| 2000 Rent Anchor | Median Gross Rent | U.S. Census Bureau, 2000 AHS | [census.gov/housing](https://www.census.gov/housing) |

---

*Report compiled with R `r paste(R.version$major, R.version$minor, sep=".")` · tidyverse · patchwork · scales.*
*All source data files are included in `data/` for full reproducibility.*
